Reinforcement \change{L}earning for \change{O}ptions\\on \change{T}arget \change{V}olatility \change{F}unds\thanks{Supported by ISCRA from Cineca, \email{iscra@cineca.it}.}

### 2023-02-18 tex/m2ef/daluiso_nastasi_pallavicini_polo_rl_for_options_on_tvf.tex
implement other referee's comments

\title{Reinforcement \change{L}earning for \change{O}ptions\\on \change{T}arget \change{V}olatility \change{F}unds\thanks{Supported by ISCRA from Cineca, \email{iscra@cineca.it}.}}

@@

A part of this work has been developed during the Master thesis of one of us (SP), where we had the opportunity to collaborate with Marco Bianchetti and Diego Pierluigi Giovannini from Intesa Sanpaolo Milan, that we thank. Moreover we wish to thank the Italian computing centre Cineca, which approved our ISCRA C project and provided us with the high-computing resources of Marconi100 for the numerical simulations of this contribution.
\\

The opinions here expressed are solely those of the authors and do not represent in any way those of their employers.

@@

Then, we introduce the strategy price process $V_t$, and we define the deflated gain process $\Bar{G}_t^V$ as given by
\begin{equation}
    \Bar{G}_t^V \coloneqq \frac{V_t}{B_t} + \int_0^t\frac{V_u\phi_u}{B_u}du \; ,
\end{equation}
where $\phi_t$ are the running fees earned by the fund manager for his activity. We assume that the strategy is self-financing, so that we can write
\begin{equation}
    d\Bar{G}_t^V = q_t \cdot d\bar{G}_t \; ,
\label{eq:self_financing}\end{equation}
where $q_t^i$ is the quantity invested in the $i$-th security.\footnote{In all formulae we use dot notation for scalar product between vectors, i.e. $a \cdot b = \sum_i a_i b_i$, or between matrix and vector, i.e. $A \cdot b = \sum_j a_{ij}b_j$ or $b \cdot A = \sum_i b_i a_{ij}$.}

@@

We highlight that, to derive the dynamics expressed in Equation~\eqref{eq:TVS_last}, we have not made any assumptions on the risky allocation strategy; thus all this argument is valid for any constraints on the process $\alpha_t$.

\section{Derivative pricing}\label{sec:Derivative}
In this section, we analyze derivative contracts linked to the TVS described previously. In particular we will focus on European style options, and we will show that under appropriate assumptions it is possible to find a closed form solution for the optimal allocation strategy which maximizes the contract price.

@@

In Figure~\ref{fig:strategy_price_comparison} we provide a comparison among the option price obtained with the optimal strategy (BS$^*$) of Equation~\eqref{eq:BS_optimal_strategy} and those with \change{the below} \remove{other} intuitive strategies (S$_A$, S$_B$, S$_C$). Here we consider the case of an at-the-money call option with spot $I_0 = 1\, \text{EUR}$, maturity $T=5 \, \text{yr}$, target volatility $\bar{\sigma}=5\%$, and a nonnegative constraint on $\alpha$. The intuitive strategies are
\begin{itemize}
\item S$_A$: invest all in the asset with the maximum forward curve at maturity;
\item S$_B$: for each market pillar before maturity invest all in the asset with minimum $\mu(t)$; 
\item S$_C$: for each market pillar before maturity invest all in the asset with minimum $\mu(t)/\|\nu(t)\|$.
\end{itemize}
As the reader can observe, our strategy outperforms any other intuitive one a practitioner might adopt.
\begin{figure}[t]
	\centering
	\includegraphics[width=4in]{strategy_price_comparison_grey.eps}
	\caption{Comparison of plain vanilla prices on the TVS by adopting different allocation strategies: the optimal Black and Scholes solution (BS$^*$) of Equation~\eqref{eq:BS_optimal_strategy} and other intuitive strategies (S$_A$, S$_B$, S$_C$)}
	\label{fig:strategy_price_comparison}
\end{figure}

@@

@@

In full generality, we assume that the time evolution of the Markovian factors governing the problem dynamics is given by  a stochastic multidimensional process in $\mathbb{R}^n$, $X_t$, which is the unique strong solution to the following It$\hat{\text{o}}$ SDE 
\begin{equation}
		dX_t = M(X_t)dt + \Sigma(X_t) \cdot dW_t \; ,
\end{equation}
where $W_t$ is a $n$-dimensional Wiener process with independent components. We point out to the reader that with this notation we are including general dynamics models like those with stochastic drift $M(X_t)$ and stochastic diffusive $\Sigma(X_t)$ coefficients.

@@

In this framework, the TVS price process dynamics is given by the SDE 
\begin{equation}
	\frac{dI_t}{I_t} = \left(r(X_t) -  \frac{\bar{\sigma} \alpha_t}{\norm{\alpha_t \cdot \nu(X_t)}}  \cdot \mu(X_t)    \right)dt +\frac{\bar{\sigma}\alpha_t }{\norm{ \alpha_t \cdot \nu(X_t) }} \cdot \nu(X_t) \cdot dW_t \; ,
\end{equation}
where the expression of $\mu(X_t)$ and $\nu(X_t)$ can be recovered by applying the It$\hat{\text{o}}$ formula to Equation~\eqref{eq:S_f}.

@@

If we assume a time-dependent BS dynamics for the risky equities ($\mu_t$, $r_t$ and $\nu_t$ deterministic), then $V = V(t,I)$ and all the derivatives w.r.t. $X$ are zero. Therefore the reduced HJB equation is
\begin{equation}
	\frac{\partial V}{\partial t} +  r(t) I \frac{\partial V}{\partial I}  + \frac{1}{2}\bar{\sigma}^2I^2\frac{\partial^2V}{\partial I^2} + \bar{\sigma}I\max_\alpha \left\{ -\frac{\partial V}{\partial I}   \frac{\alpha \cdot \mu(t)}{\norm{\alpha \cdot \nu(t)}}\right\} = 0 \; ;
\end{equation}
if the payoff is non-decreasing in $I$ by homogeneity of the SDE we get that $V$ is non-decreasing\change{.}\remove{; t}\change{T}hus the solution is given by
\begin{equation}
 \alpha^*(t) = \argmin_\alpha \frac{\alpha \cdot \mu(t)}{\|\alpha \cdot \nu(t)\|} \; ,
	\end{equation}
which is the same result expressed in Equation~\eqref{eq:BS_optimal_strategy}. On the other hand, if the payoff is non-increasing in $I$ then the solution will be the $\argmax$.

@@

\remove{Reinforcement Learning} \change{RL} is a branch of Machine Learning which allows an artificial agent to interact with an environment through actions and observations in order to maximize some notion of cumulative reward \change{called value function}. In RL\change{,} the agent is not told which actions to take but instead, it \remove{must} \change{has to} discover by trial and error which are the behaviours  yielding the highest reward. This is obtained by updating the agent policy $\pi$ which is a mapping from the environment states to the set of actions. Thus RL is independent of pre-collected data as opposed to other Machine Learning techniques. Because of its nature, RL has been successful in quantitative finance for solving control problems; among the most important RL applications in this field, we refer to \cite{Deng2017} as the pioneers in studying self-taught reinforcement trading problems, while to \cite{Halperin2020} and \cite{Kolm2019} for hedging derivatives with RL under market frictions.

@@

\subsection{Direct \change{P}olicy \change{A}pproach}\label{sec:direct_policy}
We consider an episode $\tau$ of length $m+1$ that takes place on a discrete time-grid of fixing times expressed in year fractions $\mathcal{T}\coloneqq \{T_0,...,T_k,...,T_m\}$ with $T_0\coloneqq 0$ and $T_m \coloneqq T$ maturity of the option. At a given episodic time $T_k$ the RL agent interacts with the environment: it receives a representation of the environment called state $s_k$ and on the basis of that it selects an action $a_k$ sampling from the current policy $\pi^{\theta_j}$. Here with $\theta_j$ we refer to the set of parameters through which we parameterize the agent policy at the $j$-th algorithm iteration. In our case of study the agent can choose the composition of the risky asset portfolio, so that the policy is the allocation strategy $\alpha$ introduced in Equation~\eqref{eq:TVS_elegant}:
\begin{equation}
 	a_k = \alpha_{T_k} \quad \forall T_k \in \mathcal{T} \; , \quad a_k \in \mathcal{A}\subset \mathbb{R}^n \; .
\label{eq:action}\end{equation}

@@

In our work, we have defined two different reward functions with the purpose to analyze which could help the agent to learn more efficiently the optimal policy $\pi^* = \pi^{\theta^*}$. 
Our first definition is
 \begin{equation}
 	r_{k+1}=\left\{\begin{array}{ll}
 		 D(0,T;\zeta)(I_T(\alpha(\theta_j)) - K)^+ & \text { if } T_{k+1} = T \\
 		0 & \text { otherwise }
 	\end{array}\right.
 \label{eq:reward1} \; . \end{equation}
Therefore, during the whole episode, the agent receives a nil reward except at maturity when the reward coincides with the option intrinsic value. This choice may seem too daring because the agent receives a real feedback of its actions only at the end of the episode, increasing the probability to obtain a slow learning. However, if the agent has learnt $\pi^*$, the average cumulative reward per episode will coincide with the optimal TVO price.

@@

We try to investigate the following hyper-parameters: the FFNN architecture, in particular which feature between the depth and the width of the network is more important, the activation function and the learning rate of the optimizer. Moreover, we compare the performances of two well-known optimizers in Machine Learning literature: Nadam \cite{Dozat2016} and RMSprop \cite{Hiton2012}. Since we deal with a free allocation strategy $\alpha$ that can assume negative values, we have chosen among the wide variety of activation functions the tanh and the elu. In this way, we can analyze the performance of a saturating activation function and a non-saturating one. Firstly we have performed a grid search on the learning rate starting value and we have found $10^{-3}$ a good choice in terms of speed of learning and avoiding over-fitting.
 \begin{figure}[h!]
	\centering
	\includegraphics[width=3.4in]{BS_fine_tuning_Nadam_tanh_grey.eps}
	\includegraphics[width=3.4in]{BS_fine_tuning_Nadam_elu_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the Black and Scholes \change{scenario}. The solid lines are the loss functions defined in Equation~\eqref{eq:direct_policy_loss} in function of the number of training epochs. Each continuous line represents a different neural network architecture with tanh activation function (top) and elu activation function (bottom). The optimizer adopted is the Nadam with a learning rate of $10^{-3}$. The horizontal grey-dashed line is the conservative option price obtained by maximizing the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_bs_nadam}
\end{figure}  
 \begin{figure}[h!]
	\centering
	\includegraphics[width=3.4in]{BS_fine_tuning_RMSprop_tanh_grey.eps}
	\includegraphics[width=3.4in]{BS_fine_tuning_RMSprop_elu_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the Black and Scholes \change{scenario}. The solid lines are the loss values defined in Equation~\eqref{eq:direct_policy_loss} in function of the number of training epochs. Each continuous line represents a different neural network architecture with tanh activation function (top) and elu activation function (bottom). The optimizer adopted is the RMSprop with a learning rate of $10^{-3}$. The horizontal grey-dashed line is the conservative option price obtained by maximizing the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_bs_rmsprop}
\end{figure}

@@

We take the $20\times 15 \times 5$ network with tanh from RMSprop as the best optimized network, and we run a Monte Carlo simulation with $10^6$ never-seen scenarios to check if the agent overfits the new data. 
\begin{table}
\begin{center}
\caption{Comparison of TVO prices under Black and Scholes model: analytical solution price and direct policy reinforcement learning. The option parameters are: $I_0=K=1\,[\text{EUR}]$, $T=2\,[\text{yr}]$ and $\bar{\sigma}=5\%$}
\label{tab:mc_results_bs}
\begin{tabular}{ll}
\hline\noalign{\smallskip}
Pricing method & TVO price [EUR] \\
\noalign{\smallskip}
\hline
\noalign{\smallskip}
Analytical Solution & $4.2634\times 10^{-2}$ \\
Direct policy RL & $(4.2624 \pm 0.005) \times 10^{-2}$ \\ 
\hline
\end{tabular}
\end{center}
\end{table}
\setlength{\tabcolsep}{1.4pt}
We report the results in Table~\ref{tab:mc_results_bs}: the RL price is compatible with the closed formula price. Thus the RL agent did not overfit the data during the training phase.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=4in]{LV_direct_policy_approach_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the local volatility model. The solid line with triangles is the learning curve of an agent with free actions, while the solid one is the learning curve of an agent whose actions are parameterized with the maximum drift strategy. The two agents are parameterized by a $20 \times 10 \times 5$ FFNN with tanh activation function. The loss function is computed according to Equation~\eqref{eq:direct_policy_loss}. The optimizer adopted is the RMSprop with a learning rate of $10^{-3}$ for the solid line with triangles while $10^{-4}$ for the solid one. The horizontal grey-dashed lines delimit the 99\% confidence interval of the MC price obtained by maximizing path-wise the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_lv_direct_policy}
\end{figure}  
Our first way to tackle the problem is to exploit the results obtained in the BS case. Thus we train by direct policy algorithm a deep $20\times 10\times 5$ neural network with tanh activation function and a RMSprop optimizer. We report in Figure~\ref{fig:result_lv_direct_policy} the corresponding learning curve (solid line with triangles). From this, we observe that the network has learnt some good policy since the curve grows as the number of training epochs increases until it saturates at a certain value. To try to measure the performance of the policy selected by the agent, we can build a n\"aif strategy called \quotations{baseline}. By looking at the Markovian projection in \eqref{eq:markovian_projection}-\eqref{eq:local_drift} and the first term in the HJB \eqref{eq:HJB_general}, a natural choice for the baseline is to maximize the TVS local drift. In other words this coincides by applying the path-wise the Black and Scholes solution of Equation~\eqref{eq:BS_optimal_strategy}; since here we deal with a free allocation strategy, we can simply use our analytical result \eqref{eq:analytical_solution_free_strategy}. In the same Figure~\ref{fig:result_lv_direct_policy} we report the 99\% confidence interval of the MC price obtained with the maximum drift baseline as two red-dashed lines. As we can see, the optimized loss function is compatible with the baseline price. Following the theoretical result of the HJB equation, we can assert that the agent has learnt a sub-optimal policy. Since we LV model differs from the BS one for a corrective term in the HJB \eqref{eq:HJB_general}, we expect that the optimal solution  in the LV framework will be in a close region of the maximum drift strategy. Thus we train another agent with the direct policy learning by parameterizing its action with the baseline strategy and choosing a smaller learning rate of $10^{-4}$. With this parameterization, at each observational time $T_k\in\mathcal{T}$ the risky allocation strategy is obtained by summing the network output with the Equation~\eqref{eq:analytical_solution_free_strategy}. Even in this case, the corresponding learning curve (solid black line in Figure~\ref{fig:result_lv_direct_policy}) is stuck in the maximum drift strategy. The first possible reason for this behaviour is that the learning strategy of the direct policy approach is too simple for the learning task. The second interpretation is that the volatility versor in the HJB does not affect significantly the optimum location since it is a second-order term and is lost in the MC error.

@@

Gatheral, J.: The volatility surface: a practitioner's guide. Wiley Finance (2006)

>>>>

\title{Reinforcement \change{L}earning for \change{O}ptions\\on \change{T}arget \change{V}olatility \change{F}unds\thanks{\change{Our sincere thanks go to Marco Bianchetti and Diego Pierluigi Giovannini of Intesa Sanpaolo Milan, as well as to the Italian computing center Cineca, who generously supported our numerical simulations through an ISCRA C grant. The opinions here expressed are solely those of the authors and do not represent in any way those of their employers.}}}

@@

\remove{A part of this work has been developed during the Master thesis of one of us (SP), where we had the opportunity to collaborate with Marco Bianchetti and Diego Pierluigi Giovannini from Intesa Sanpaolo Milan, that we thank. Moreover we wish to thank the Italian computing centre Cineca, which approved our ISCRA C project and provided us with the high-computing resources of Marconi100 for the numerical simulations of this contribution.
\\

The opinions here expressed are solely those of the authors and do not represent in any way those of their employers.}

@@

Then, we introduce the strategy price process $V_t$, and we define the deflated gain process $\Bar{G}_t^V$ as \remove{given by}
\begin{equation}
    \Bar{G}_t^V \coloneqq \frac{V_t}{B_t} + \int_0^t\frac{V_u\phi_u}{B_u}du \; ,
\end{equation}
where $\phi_t$ are the running fees earned by the fund manager for his activity. We assume that the strategy is self-financing, so that we can write
\begin{equation}
    d\Bar{G}_t^V = q_t \cdot d\bar{G}_t \; ,
\label{eq:self_financing}\end{equation}
where $q_t^i$ is the quantity invested in the $i$-th security.\footnote{In all formulae we use dot notation for scalar product between vectors, i.e. $a \cdot b = \sum_i a_i b_i$, or between matrix and vector, i.e. $A \cdot b = \sum_j a_{ij}b_j$ or $b \cdot A = \sum_i b_i a_{ij}$.}

@@

We highlight that, to derive the dynamics expressed in Equation~\eqref{eq:TVS_last}, we have not made any assumptions on the risky allocation strategy; thus all this argument is valid for any constraints on the process $\alpha_t$\change{ as, for instance, the case of contracts in which the fund manager is restricted to holding only long positions meaning that $\alpha_t^i>0$ $\forall i = 1,\dots,n$, or situations where the combined total of long and short trades must be capped at a specific contractual amount}.

\section{Derivative pricing}\label{sec:Derivative}
In this section, we analyze derivative contracts linked to the TVS described previously. In particular\change{,} we will focus on European style options, and we will show that under appropriate assumptions it is possible to find a closed form solution for the optimal allocation strategy which maximizes the contract price.

@@

In Figure~\ref{fig:strategy_price_comparison} we provide a comparison among the option price obtained with the optimal strategy (BS$^*$) of Equation~\eqref{eq:BS_optimal_strategy} and those with \change{the below expert-based single-asset} \remove{other intuitive} strategies (S$_A$, S$_B$, S$_C$). Here we consider the case of an at-the-money call option with spot $I_0 = 1\, \text{EUR}$, maturity $T=5 \, \text{yr}$, target volatility $\bar{\sigma}=5\%$, and a nonnegative constraint on $\alpha$. The intuitive strategies are
\begin{itemize}
\item S$_A$: invest all in the asset with the maximum forward curve at maturity; \change{this strategy constists of selecting a static portfolio that is never rebalanced, such that the TVO moneyness is maximized;}
\item S$_B$: \remove{for each market pillar before maturity} \change{$\forall t \in \mathcal{T}$} invest all in the asset with minimum $\mu(t)$; \change{in this case we consider a fund manager who rebalances the weights $\alpha(t)$ over the life of the portfolio by selecting the asset with the highest expected return at a given point in time;} 
\item S$_C$: \remove{for each market pillar before maturity} \change{$\forall t \in \mathcal{T}$} invest all in the asset with minimum $\mu(t)/\|\nu(t)\|$\remove{.}\change{; the strategy is similar to S$_B$, except in this case, the manager maximizes the expected return weighted for the target volatility mechanism.}
\end{itemize}
As the reader can observe, our strategy outperforms any other intuitive one a practitioner might adopt.
\begin{figure}[t]
	\centering
	\includegraphics[width=4in]{strategy_price_comparison_grey.eps}
	\caption{Comparison of plain vanilla prices on the TVS by adopting different allocation strategies: the optimal Black and Scholes solution (BS$^*$) of Equation~\eqref{eq:BS_optimal_strategy} and \change{three expert-based single-asset} strategies (S$_A$, S$_B$, S$_C$)}
	\label{fig:strategy_price_comparison}
\end{figure}

\change{In the next subsection we derive the analytical solution of the problem \eqref{eq:BS_optimal_strategy} under the assumption of unconstrained allocation strategy $\alpha_t$}

@@

\change{In the following subsection we prove that an analytical solution of \eqref{eq:BS_optimal_strategy} exists also assuming a nonnegative constrained allocation strategy. Under this assumption we show that the optimal strategy is a single-asset one called bang bang solution.}

@@

In full generality, we assume that the time evolution of the Markovian factors governing the problem dynamics is given by  a stochastic multidimensional process in $\mathbb{R}^n$, $X_t$, which is the unique strong solution to the following It$\hat{\text{o}}$ \remove{SDE} \change{stochastic differential equation} 
\begin{equation}
		dX_t = M(X_t)dt + \Sigma(X_t) \cdot dW_t \; ,
\end{equation}
where $W_t$ is a $n$-dimensional Wiener process with independent components. We point out to the reader that with this notation we are including general dynamics models like those with stochastic drift $M(X_t)$ and stochastic diffusive $\Sigma(X_t)$ coefficients.

@@

In this framework, the TVS price process dynamics is given by the \remove{SDE} \change{stochastic differential equation} 
\begin{equation}
	\frac{dI_t}{I_t} = \left(r(X_t) -  \frac{\bar{\sigma} \alpha_t}{\norm{\alpha_t \cdot \nu(X_t)}}  \cdot \mu(X_t)    \right)dt +\frac{\bar{\sigma}\alpha_t }{\norm{ \alpha_t \cdot \nu(X_t) }} \cdot \nu(X_t) \cdot dW_t \; ,
\end{equation}
where the expression of $\mu(X_t)$ and $\nu(X_t)$ can be recovered by applying the It$\hat{\text{o}}$ formula to Equation~\eqref{eq:S_f}.

@@

If we assume a time-dependent BS dynamics for the risky equities ($\mu_t$, $r_t$ and $\nu_t$ deterministic), then $V = V(t,I)$ and all the derivatives w.r.t. $X$ are zero. Therefore the reduced HJB equation is
\begin{equation}
	\frac{\partial V}{\partial t} +  r(t) I \frac{\partial V}{\partial I}  + \frac{1}{2}\bar{\sigma}^2I^2\frac{\partial^2V}{\partial I^2} + \bar{\sigma}I\max_\alpha \left\{ -\frac{\partial V}{\partial I}   \frac{\alpha \cdot \mu(t)}{\norm{\alpha \cdot \nu(t)}}\right\} = 0 \; ;
\end{equation}
if the payoff is non-decreasing in $I$\change{,} by homogeneity of the \remove{SDE} \change{stochastic differential equation} we get that $V$ is non-decreasing\change{.}\remove{; t}\change{T}hus the solution is given by
\begin{equation}
 \alpha^*(t) = \argmin_\alpha \frac{\alpha \cdot \mu(t)}{\|\alpha \cdot \nu(t)\|} \; ,
	\end{equation}
which is the same result expressed in Equation~\eqref{eq:BS_optimal_strategy}. On the other hand, if the payoff is non-increasing in $I$ then the solution will be the $\argmax$.

@@

\remove{Reinforcement Learning} \change{RL} is a branch of Machine Learning which allows an artificial agent to interact with an environment through actions and observations in order to maximize some notion of cumulative reward \change{called value function}. In RL\change{,} the agent is not told which actions to take but instead, it \remove{must} \change{has to} discover by trial and error which are the behaviours  yielding the highest reward. This is obtained by updating the agent policy $\pi$ which is a mapping from the environment states to the set of actions. Thus RL is independent of pre-collected data as opposed to other Machine Learning techniques\change{, that have to be fed a pre-existing dataset to learn from}. Because of its nature, RL has been successful in quantitative finance for solving control problems; among the most important RL applications in this field, we refer to \cite{Deng2017} as the pioneers in studying self-taught reinforcement trading problems, while to \cite{Halperin2020} and \cite{Kolm2019} for hedging derivatives with RL under market frictions.

@@

\subsection{Direct \change{P}olicy \change{A}pproach}\label{sec:direct_policy}
\change{In RL, the learning phase takes place within a set of so-called episodes, which refer to a single run or instance of interaction between the RL agent and an environment, where the agent takes a sequence of actions to achieve a specific goal.}

\change{For our problem, we} \remove{We} consider an episode $\tau$ of length $m+1$ that \remove{takes place on} \change{unfolds over} a discrete time-grid of fixing times expressed in year fractions $\mathcal{T}\coloneqq \{T_0,...,T_k,...,T_m\}$ with $T_0\coloneqq 0$ and $T_m \coloneqq T$ maturity of the option. At a given episodic time $T_k$ the RL agent interacts with the environment: it receives a representation of the environment called state $s_k$ and on the basis of that it selects an action $a_k$ sampling from the current policy $\pi^{\theta_j}$. Here with $\theta_j$ we refer to the set of parameters through which we parameterize the agent policy at the $j$-th algorithm iteration. In our case of study\change{,} the agent\change{, representing the fund manager, has the right to select} \remove{can choose} the composition of the risky asset portfolio, so that the policy is the allocation strategy $\alpha$ introduced in Equation~\eqref{eq:TVS_elegant}:
\begin{equation}
 	a_k = \alpha_{T_k} \quad \forall T_k \in \mathcal{T} \; , \quad a_k \in \mathcal{A}\subset \mathbb{R}^n \; .
\label{eq:action}\end{equation}

@@

In our work, we have defined two different reward functions with the purpose to analyze which could help the agent to learn more efficiently the optimal policy $\pi^* = \pi^{\theta^*}$. 
Our first definition is
 \begin{equation}
 	r_{k+1}\change{\coloneqq}\left\{\begin{array}{ll}
 		 D(0,T;\zeta)(I_T(\alpha(\theta_j)) - K)^+ & \text { if } T_{k+1} = T \\
 		0 & \text { otherwise }
 	\end{array}\right.
 \label{eq:reward1} \; . \end{equation}
Therefore, during the whole episode, the agent receives a nil reward except at maturity when the reward coincides with the option intrinsic value. This choice may seem too daring because the agent receives a real feedback of its actions only at the end of the episode, increasing the probability to obtain a slow learning. However, if the agent has learnt $\pi^*$, the average cumulative reward per episode will coincide with the optimal TVO price.

@@

We try to investigate the following hyper-parameters: the FFNN architecture, in particular which feature between the depth and the width of the network is more important, the activation function and the learning rate of the optimizer. Moreover, we compare the performances of two well-known optimizers in Machine Learning literature: Nadam \cite{Dozat2016} and RMSprop \cite{Hiton2012}. Since we deal with a free allocation strategy $\alpha$ that can assume negative values, we have chosen among the wide variety of activation functions \change{\cite{Goodfellow2016}} the tanh and the elu. In this way, we can analyze the performance of a saturating activation function and a non-saturating one. Firstly we have performed a grid search on the learning rate starting value and we have found $10^{-3}$ a good choice in terms of speed of learning and avoiding over-fitting.
 \begin{figure}[h!]
	\centering
	\includegraphics[width=3.4in]{BS_fine_tuning_Nadam_tanh_grey.eps}
	\includegraphics[width=3.4in]{BS_fine_tuning_Nadam_elu_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the Black and Scholes \change{scenario}. The solid lines are the loss functions defined in Equation~\eqref{eq:direct_policy_loss} in function of the number of training epochs. Each continuous line represents a different neural network architecture with tanh activation function (top) and elu activation function (bottom). The optimizer adopted is the Nadam with a learning rate of $10^{-3}$. The horizontal grey-dashed line is the conservative option price obtained by maximizing the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_bs_nadam}
\end{figure}  
 \begin{figure}[h!]
	\centering
	\includegraphics[width=3.4in]{BS_fine_tuning_RMSprop_tanh_grey.eps}
	\includegraphics[width=3.4in]{BS_fine_tuning_RMSprop_elu_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the Black and Scholes \change{scenario}. The solid lines are the loss values defined in Equation~\eqref{eq:direct_policy_loss} in function of the number of training epochs. Each continuous line represents a different neural network architecture with tanh activation function (top) and elu activation function (bottom). The optimizer adopted is the RMSprop with a learning rate of $10^{-3}$. The horizontal grey-dashed line is the conservative option price obtained by maximizing the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_bs_rmsprop}
\end{figure}

@@

We take the $20\times 15 \times 5$ network with tanh from RMSprop as the best optimized network, and we run a Monte Carlo simulation with $10^6$ never-seen scenarios to check if the agent overfits the new data. 
\begin{table}
\begin{center}
\caption{Comparison of TVO prices under Black and Scholes \change{scenario}: analytical solution price and direct policy reinforcement learning. The option parameters are: $I_0=K=1\,[\text{EUR}]$, $T=2\,[\text{yr}]$ and $\bar{\sigma}=5\%$}
\label{tab:mc_results_bs}
\begin{tabular}{ll}
\hline\noalign{\smallskip}
Pricing method & TVO price [EUR] \\
\noalign{\smallskip}
\hline
\noalign{\smallskip}
Analytical Solution & $4.2634\times 10^{-2}$ \\
Direct policy RL & $(4.2624 \pm 0.005) \times 10^{-2}$ \\ 
\hline
\end{tabular}
\end{center}
\end{table}
\setlength{\tabcolsep}{1.4pt}
We report the results in Table~\ref{tab:mc_results_bs}: the RL price is compatible with the closed formula price. Thus the RL agent did not overfit the data during the training phase.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=4in]{LV_direct_policy_approach_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the local volatility \change{scenario}. The solid line with triangles is the learning curve of an agent with free actions, while the solid one is the learning curve of an agent whose actions are parameterized with the maximum drift strategy. The two agents are parameterized by a $20 \times 10 \times 5$ FFNN with tanh activation function. The loss function is computed according to Equation~\eqref{eq:direct_policy_loss}. The optimizer adopted is the RMSprop with a learning rate of $10^{-3}$ for the solid line with triangles while $10^{-4}$ for the solid one. The horizontal grey-dashed lines delimit the 99\% confidence interval of the MC price obtained by maximizing path-wise the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_lv_direct_policy}
\end{figure}  
Our first way to tackle the problem is to exploit the results obtained in the BS case. Thus we train by direct policy algorithm a deep $20\times 10\times 5$ neural network with tanh activation function and a RMSprop optimizer. We report in Figure~\ref{fig:result_lv_direct_policy} the corresponding learning curve (solid line with triangles). From this, we observe that the network has learnt some good policy since the curve grows as the number of training epochs increases until it saturates at a certain value. To try to measure the performance of the policy selected by the agent, we can build a n\"aif strategy called \quotations{baseline}. By looking at the Markovian projection in \eqref{eq:markovian_projection}-\eqref{eq:local_drift} and the first term in the HJB \eqref{eq:HJB_general}, a natural choice for the baseline is to maximize the TVS local drift. In other words this coincides by applying the path-wise the Black and Scholes solution of Equation~\eqref{eq:BS_optimal_strategy}; since here we deal with a free allocation strategy, we can simply use our analytical result \eqref{eq:analytical_solution_free_strategy}. In the same Figure~\ref{fig:result_lv_direct_policy} we report the 99\% confidence interval of the MC price obtained with the maximum drift baseline as two red-dashed lines. As we can see, the optimized loss function is compatible with the baseline price. Following the theoretical result of the HJB equation, we can assert that the agent has learnt a sub-optimal policy. Since we LV model differs from the BS one for a corrective term in the HJB \eqref{eq:HJB_general}, we expect that the optimal solution  in the LV framework will be in a close region of the maximum drift strategy. Thus we train another agent with the direct policy learning by parameterizing its action with the baseline strategy and choosing a smaller learning rate of $10^{-4}$. With this parameterization, at each observational time $T_k\in\mathcal{T}$ the risky allocation strategy is obtained by summing the network output with the Equation~\eqref{eq:analytical_solution_free_strategy}. Even in this case, the corresponding learning curve (solid black line in Figure~\ref{fig:result_lv_direct_policy}) is stuck in the maximum drift strategy. The first possible reason for this behaviour is that the learning strategy of the direct policy approach is too simple for the learning task. The second interpretation is that the volatility versor in the HJB does not affect significantly the optimum location since it is a second-order term and is lost in the MC error.

@@

\change{Gatheral, J.: The volatility surface: a practitioner's guide. Wiley Finance (2006)}
\bibitem{Goodfellow2016}
\change{Goodfellow, I., Bengio, Y., Courville, A.: Deep learning. MIT Press (2016)}
